DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-05-26 manuscript/figures-alt.tex
cross validation XY figure
addressing topologies comment

>>>>

\begin{figure}[!h]
\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
\caption{Topologies obtained from OpenTree's synthetic phylogeny. A) Topology od 2333 tips, encompassing bird species within the Fringillidae following the NCBI taxonomy (black) as well as all other bird species that share the same mrca node in the synthetic tree (purple). B) Topology encompassing bird species within the Fringillidae only, resulting from pruning purple branches from topology A.
Bird species belonging to the family Fringillidae do not form a monophyletic group.
(Alström et al. \protect\hyperlink{ref-}{2014},
Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
Price et al. \protect\hyperlink{ref-price2014niche}{2014},
Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012})
}
\label{fig:fringillidae-topologies}
\end{figure}

### 2022-05-26 manuscript/manuscript.Rmd
addressing topologies comment

nocite: |
  @barker2013going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---

@@

Here we present the DateLife project and its core software application, available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
The software features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem], data stored and available in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare the data.

@@

_**Summarizing search results.-**_
The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. As explained in the "Description" section, a tree topology to summarize age data upon is required.
By default, `datelife` uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, Fringillidae is not a monophyletic family
[@burns2014phylogenetics;
@burleigh2015building;
@Jetz2012;
@lerner2011multilocus;
@lovette2010comprehensive;
@pulgarin2013multilocus;
@chaves2013biogeography;
@lamichhaney2015evolution;
@parchman2007coevolution;
@selvatti2015paleogene;
@moyle2016tectonic;
@odeen2011evolution;
@claramunt2015new;
@beresford2005african;
@kimball2019phylogenomic;
@barker2015new;
@klicka2014comprehensive;
@barker2013going;
@treplin2008molecular;
@price2014niche;
@hackett2008phylogenomic;
@barker2014mitogenomic;
@gibb2015new;
@barker2004phylogeny;
@zuccon2012phylogenetic;
@oliveros2019earth;
@johansson2008phylogenetic;
@alstrom2014discovery;
@powell2014comprehensive;
@bryson2014diversification;
@packert2012horizontal;
@tietze2013complete].

Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.

>>>>

nocite: |
  @barker2013going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche,
  @alstrom2014discovery,
  @barker2004phylogeny,
  @barker2013going,
  @barker2014mitogenomic,
  @barker2015new,
  @beresford2005african,
  @bryson2014diversification,
  @burleigh2015building,
  @burns2014phylogenetics,
  @chaves2013biogeography,
  @claramunt2015new,
  @gibb2015new,
  @hackett2008phylogenomic,
  @Jetz2012,
  @johansson2008phylogenetic,
  @kimball2019phylogenomic,
  @klicka2014comprehensive, 
  @lamichhaney2015evolution,
  @lerner2011multilocus,
  @lovette2010comprehensive,
  @moyle2016tectonic,
  @odeen2011evolution,
  @oliveros2019earth,
  @packert2012horizontal,
  @parchman2007coevolution,
  @powell2014comprehensive,
  @price2014niche,
  @pulgarin2013multilocus,
  @selvatti2015paleogene,
  @tietze2013complete,
  @treplin2008molecular,
  @zuccon2012phylogeneti,
---

@@

Here we present the DateLife project which has the main goal of capturing data from published chronograms, and making these data readily accessible to users for reuse and reanalysis.
DateLife's core software application is available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org. It features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem], data stored and available in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare the data.

@@

_**Summarizing search results.-**_
The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. As explained in the "Description" section, a tree topology to summarize age data upon is required.
By default, `datelife` uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group.
Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.

### 2022-05-26 manuscript/manuscript.tex
addressing topologies comment

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4607}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

Here we present the DateLife project and its core software application, available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
The software features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.

@@

\emph{\textbf{Summarizing search results.-}}
The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
By default, \texttt{datelife} uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, Fringillidae is not a monophyletic family
(Alström et al., 2014; Barker, 2014; Barker et al., 2013, 2015; Barker, Cibois, Schikler, Feinstein, \& Cracraft, 2004; Beresford, Barker, Ryan, \& Crowe, 2005; Bryson Jr et al., 2014; Burleigh, Kimball, \& Braun, 2015; Burns et al., 2014; Chaves, Hidalgo, \& Klicka, 2013; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hackett et al., 2008; Jetz et al., 2012; Johansson, Fjeldså, \& Bowie, 2008; Kimball et al., 2019; Klicka et al., 2014; Lamichhaney et al., 2015; Lerner, Meyer, James, Hofreiter, \& Fleischer, 2011; Lovette et al., 2010; Moyle et al., 2016; Ödeen, Håstad, \& Alström, 2011; Oliveros et al., 2019; Päckert et al., 2012; Parchman, Benkman, \& Mezquida, 2007; Powell et al., 2014; Price et al., 2014; Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka, 2013; Selvatti, Gonzaga, \& Moraes Russo, 2015; Tietze, Päckert, Martens, Lehmann, \& Sun, 2013; Treplin et al., 2008; Zuccon, Prŷs-Jones, Rasmussen, \& Ericson, 2012).

Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.

@@

At the time of writing of this manuscript (May 22, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
Barker, F. K., Burns, K. J., Klicka, J., Lanyon, S. M., \& Lovette, I. J. (2015). New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. \emph{{The Auk: Ornithological Advances}}, \emph{132}(2), 333--348.

@@

\leavevmode\vadjust pre{\hypertarget{ref-zuccon2012phylogenetic}{}}%
Zuccon, D., Prŷs-Jones, R., Rasmussen, P. C., \& Ericson, P. G. (2012). The phylogenetic relationships and generic limits of finches (fringillidae). \emph{Molecular Phylogenetics and Evolution}, \emph{62}(2), 581--596.

@@

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4586}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

Here we present the DateLife project which has the main goal of capturing data from published chronograms, and making these data readily accessible to users for reuse and reanalysis.
DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org. It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.

@@

\emph{\textbf{Summarizing search results.-}}
The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
By default, \texttt{datelife} uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group.
Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.

@@

At the time of writing of this manuscript (May 26, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
Barker, F. K., Burns, K. J., Klicka, J., Lanyon, S. M., \& Lovette, I. J. (2015). New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. \emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.

@@

@@

\begin{figure}[!h]
\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
\caption{Fringillidae topologies available from OpenTree's synthetic phylogeny. A) Topology encompassing bird species within the Fringillidae within the NCBI taxonomy (black) as well as all species that share the same mrca node in the synthetic tree (purple). B) Paraphyletic topology encompassing bird species within the Fringillidae only. It results from pruning purple branches from topology in A.
Bird species belonging to the family Fringillidae do not form a monophyletic group.
(Alström et al. \protect\hyperlink{ref-}{2014},
Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
Price et al. \protect\hyperlink{ref-price2014niche}{2014},
Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012})
}
\label{fig:fringillidae-topologies}
\end{figure}
